---
title: "NYT API Retrieval"
author: "Lucian Scher"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(jsonlite)
library(tidyverse)
library(tidytext)
library(ggplot2)
```

```{r API-key}
# Get a free key at https://developer.nytimes.com/get-started
#   usethis::edit_r_environ()
#   NYT_API_KEY=your_api_key_here
API_KEY <- Sys.getenv("NYT_API_KEY")
if (API_KEY == "") API_KEY <- Sys.getenv("NYT_KEY")

# Or paste your key here:
if (API_KEY == "") {
  API_KEY <- "your_api_key_here"
}

if (API_KEY == "" || API_KEY == "your_api_key_here") {
  stop("No API key found. Set NYT_API_KEY in .Renviron, restart R, then re-run this chunk.")
}
```

Retrieve New York Times articles for the Trump sentiment analysis project. Same timeframe as `data_exploration.Rmd`: **July 7, 2025** through **May 20, 2026**.

<https://developer.nytimes.com/>

**Do not query the full date range in one call** — that returns 10,000 hits, caps at 1,000, and hits rate limits fast. Use the 2-week window script below instead.

### Run in background (tmux) — recommended

```bash
cd Trump_sentiment_analysis
bash run_nyt.sh
tail -f nyt_fetch.log          # check progress
tmux attach -t nyt-fetch       # reattach
```

The script retries on `429` errors, waits 15s between pages, caches each page in `nyt_cache/`, and resumes where it left off if interrupted.

### Probe: total hits for full range

```{r api}
searchQ = URLencode("trump")
url = paste0(
  'https://api.nytimes.com/svc/search/v2/articlesearch.json?q=', searchQ,
  '&begin_date=20250707&end_date=20260520&api-key=', API_KEY
)

t = fromJSON(url, flatten = TRUE)
hits = t$response$meta$hits
hits
```

### Retrieve articles (2-week windows)

```{r results}
# Run API-key chunk first, then (from EDS-231-text-sentiment):
source("scripts/run_nyt_retrieval.R")
```

### Quick check

```{r check}
class(nytDat)
dim(nytDat)
names(nytDat)
head(nytDat)
```
